---
title: "Mother Wage Penalty"
output:
  pdf_document: default
  html_document: default
---

Across industrialized countries, it is a well-studied phenomenon 
that childless women are paid more on average than mothers. In 
this exercise, we use survey data to investigate and measure
the wage gap between women with children and childless women.

The exercise is based on: Wei-hsin Yu and J
anet Chen-Lan Kuo. 2017. [The Motherhood Wage Penalty by Work 
Conditions: How Do Occupational Characteristics Hinder or 
Empower Mothers?](https://doi.org/10.1177/0003122417712729)" 
*American Sociological Review* 82(4): 744-769.

In this paper, the authors examine the association between 
the so-called *mother wage penalty* (i.e., the pay gap between mothers 
and non-mothers) and occupational characteristics. Three prominent 
explanations for the motherhood wage penalty--"stressing work-family conflict 
and job performance," "compensating differentials," and "employer 
discrimination"--provide hypotheses about the relationship 
between penalty changes and occupational characteristics.
The authors use data from 16 waves of the National Longitudinal 
Survey of Youth to estimate the effects of five occupational 
characteristics on the mother wage penalty and to test these 
hypotheses. 

This paper uses a type of data known as 'panel data.' Panel data consist of 
observations on the same people over time. In this example, we are going to 
analyze the same women over multiple years. When analyzing panel data, each 
time period is referred to as a *wave*, so here each year is a wave.

The data file is `yu2017sample.csv`, which is a CSV file. The names 
and descriptions of variables are:

 -------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `PUBID`              ID of woman

 `year`               Year of observation
 
 `wage`               Hourly wage, in cents

 `numChildren`        Number of children that the woman has (in this wave)
 
 `age`                Age in years
 
 `region`             Name of region (North East = 1, North Central = 2, South = 3, West = 4)
 
 `urban`              Geographical classification (urban = 1, otherwise = 0)
      
 `marstat`            Marital status
 
 `educ`               Level of education
 
 `school`             School enrollment (enrolled = `TRUE`, otherwise = `FALSE`)
 
 `experience`         Experience since 14 years old, in days
 
 `tenure`             Current job tenure, in years
 
 `tenure2`            Current job tenure in years, squared
 
 `fullTime`           Employment status (employed full-time = `TRUE`, otherwise = `FALSE`)
 
 `firmSize`           Size of the firm
 
 `multipleLocations`  Multiple locations indicator (firm with multiple locations = 1, otherwise = 0)
 
 `unionized`          Job unionization status (job is unionized = 1, otherwise = 1)
 
 `industry`           Job's industry type
 
 `hazardous`          Hazard measure for the job (between 1 and 2)
 
 `regularity`         Regularity measure for the job (between 1 and 5)
 -------------------------------------------------------------------------------
 
## Question 1

How many different women are in the data? How many observations 
  per year? We will refer to each row as a 'person-year observation' 
  since the row contains data on a given person in a particular year.
  
  In a few sentences, describe one advantage and one disadvantage of using 
  a contemporary cohort of women rather than an older cohort in 
  estimating the predictors of the mother wage gap.

## Answer 1

```{r}
mwp <- read.csv("data/yu2017sample.csv") 
length(unique(mwp$PUBID))
table(mwp$year)
```

`r length(unique(mwp$PUBID))` different women are in the data. The trend of observations 
per year is an gradual increase chronologically, with the first year of data, 1997,
having `r table(mwp$year)[1]` person-year observations and the last year of data,
2013, having `r tail(table(mwp$year),1)` person-year observations.

Advantages include: updates the research by a generation; allows us to 
use more detailed occupational data; and mothers have a wider range of 
occupations now was not counted. 

Disadvantages include: misses the experiences
of older mothers and the full longitudinal impact of motherhood wage penalty.

## Question 2

`numChildren` is the variable representing how many children 
  the woman had at the time of an observation. Please provide a table 
  that shows the proportion of observations by the number of children.
  Provide a brief substantive interpretation of the results.

## Answer 2

```{r}
prop.table(table(mwp$numChildren))
```
`r round(prop.table(table(mwp$numChildren)), 2)[1]*100`% women at the time of an observation are childless, `r round(prop.table(table(mwp$numChildren)), 2)[2]*100`% women have one child, `r round(prop.table(table(mwp$numChildren)), 2)[3]*100`% are mothers with two children, and `r sum(round(prop.table(table(mwp$numChildren)), 2)[-c(1:3)])*100`% women have three or more children.

## Question 3

Create a new indicator variable `isMother` that takes a value of 1 
  if the woman has at least one child in that year and a value 
  of 0 otherwise. Tabulate the new variable. Briefly comment on 
  the results.

## Answer 3

```{r}
mwp$isMother <- ifelse(mwp$numChildren > 0, 1, 0) 
table(mwp$isMother)
```
`r round(table(mwp$isMother)[1]/sum(table(mwp$isMother)),2)*100`% of women observed are childless.

## Question 4

Create a new variable called `logwage` that is the log of `wage`. 
  Make two boxplots, one for `wage` and the other for `logwage`, 
  as a function of educational level (`educ`). Compare the two boxplots 
  and discuss the purpose of the log transformation. 

## Answer 4

```{r}
mwp$logwage <- log(mwp$wage) 
par(mfrow=c(1,2))
boxplot(wage ~ educ, data = mwp) 
boxplot(logwage ~ educ, data = mwp)
```

As you can see from the boxplot of wage, the wage variable is right skewed. 
Taking the log of wage reduces the skewness of the variable and makes the 
distribution approximately normal. Further, as the authors point out, taking
the log of a variable will allow us to interpret regression coefficients
as percent changes in hourly wages. Note that this interpretation is 
just an approximation (i.e., a coefficient of 0.08 is roughly a 8 percent 
increase) and that the approximation is best for small coefficients.



## Question 5

Make a simple scatterplot of `logwage`against `tenure` on the job and then do the same for `tenure2`.
What does the different between the two plots tell us?

Finally, calculate the correlation between `logwage`and `tenure` as well as the correlation between `logwage`and `tenure2`.

## Answer 5

```{r}
plot(mwp$tenure, mwp$logwage, main = " Relationship between tenure and log wages",
     xlab = "Tenure on the Job",
     ylab = "lnwage",
     col = "steelblue")

plot(mwp$tenure2, mwp$logwage, main = " Relationship between tenure2 and log wages",
     xlab = "Tenure on the Job, squared",
     ylab = "lnwage",
     col = "indianred")

cor(mwp$tenure, mwp$logwage)

cor(mwp$tenure2, mwp$logwage)

```





## Question 6

In the same graph, plot the mean `logwage` against year for mothers, 
  then for non-mothers in a different color or line type. Include 
  a legend and a proper title. Make sure the figure and axes are 
  clearly labeled. Give a brief interpretation of the results.
 
## Answer 6 
 
```{r}
wagebyyear <- tapply(mwp$logwage, INDEX = list(mwp$isMother, mwp$year), FUN = mean) 
years <- sort(unique(mwp$year))
plot(years, wagebyyear[1,], type = "l", main = "Mean Wage by Year",
     xlab = "Year",
     ylab = "Mean lnwage",
     col = "blue")
lines(years, wagebyyear[2,], type = "l", col = "red") 
legend(x = 2008, y = 6.4,
       legend = c("Mother", "Non-mother"), 
       pch = c(15, 15),
       col = c("Red", "Blue"), 
       cex = 0.6)
```

Wage increases for both mothers and non-mothers across years. 
The wage gap between mother and non-mother became evident since 2004.

## Question 7

Create a quantile-quantile plot comparing the distribution of log wages
  of mothers and non-mothers. Comment briefly on the resulting plot.
 
## Answer 7
 
```{r}
nonMotherLogWages <- mwp$logwage[mwp$isMother == 0]
motherLogWages <- mwp$logwage[mwp$isMother == 1]
qqplot(nonMotherLogWages, motherLogWages)
abline(a=0, b=1)
```

The quantile-quantile plot indicates that the distribution 
of log wages is relatively more dispersed for non-mothers than for
mothers. It also indicates that mothers on the low end of the log wage
distribution are comparatively wealthier than non-mothers on the
low end of the log wage distribution and that mothers on the high
end of the log wage distribution are comparatively wealther than 
non-mothers on the high end of the log wage distribution.
